Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated

### 2017-02-22 ms_pre_post.Rmd
Finish edits through the Introduction

@@

A recent genetically-explicit individual-based model examined the conditions under which polyandry is predicted to evolve solely due to selection stemming from pre-copulatory inbreeding avoidance [@Duthie]. Evolution of polyandry occurred only under highly restricted conditions, requiring that direct negative selection (i.e., 'costs') on polyandry were small, that very few males were available for a female's initial mate choice but many were available for additional mate choice(s), or that polyandry was conditionally expressed when a focal female was related to her initial mate [@Duthie]. Without these conditions, increasing polyandry tended to increase rather than decrease the overall degree of inbreeding that females experienced, and hence reduced offspring fitness. This occurred because, once pre-copulatory inbreeding avoidance evolved, polyandrous females had already chosen available unrelated males as their initial mates. Their additional mates, chosen from the remaining male population, were therefore increasingly likely to include relatives. Evolution of polyandry purely to facilitate pre-copulatory inbreeding avoidance, and hence the resulting degree of inbreeding avoidance, was consequently restricted [@Duthie].

However, if post-copulatory inbreeding avoidance could also evolve, polyandrous females could further bias paternity towards unrelated males within their set of chosen mates. Post-copulatory inbreeding avoidance might consequently reduce the cost of polyandry stemming from the accumulation of related mates, potentially facilitating evolution of polyandry to avoid inbreeding under broader conditions, and driving further consequent evolution of pre-copulatory or post-copulatory mate choice strategies. However, once both pre-copulatory and post-copulatory inbreeding avoidance, and resulting polyandry, can all evolve, the long-term dynamics of these reproductive strategies become difficult to conceptualise and predict. Strong inbreeding depression might drive initial evolution of both pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. However, if evolution of polyandry and post-copulatory inbreeding avoidance renders pre-copulatory inbreeding avoidance redundant, or vice versa, then only one inbreeding avoidance strategy might be maintained in the long-term.

The few previous models that considered the evolution of inbreeding avoidance through mate choice (as opposed to dispersal) have implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models are useful for isolating the invasion fitness of this single strategy. However, when both pre-copulatory and post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will simultaneously invade a randomly mating population, nor that the invasion fitness of one strategy will be independent of the existence of invasion fitness of the other strategy. For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then new alleles underlying polyandry and post-copulatory inbreeding avoidance might be precluded from invading a population because the phenotypic effect of such alleles on the degree of inbreeding could be negligible. Conversely, fixation of alleles underlying polyandry and post-copulatory inbreeding avoidance might eliminate positive selection on alleles underlying pre-copulatory inbreeding avoidance. New theory, guided by modelling that evaluates invasion dynamics given appropriate sets of evolutionary constraints, is therefore needed.

Ultimately, the absolute and relative frequencies of alleles underlying pre-copulatory and post-copulatory inbreeding avoidance and polyandry will depend not only on the magnitudes of positive indirect selection stemming from reduced inbreeding depression in females' offspring, but also on the magnitudes of direct negative selection on resulting phenotypes (i.e., the direct fitness costs of expressing each reproductive strategy). Polyandrous females can pay costs of mate searching or mating, for example, because these activities increase predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice and consequent reluctance to mate with relatives might pay costs of increased risk of complete mating failure or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay energetic costs associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs of pre-copulatory and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly strategy might become fixed over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy. However, such trajectories will also depend on the indirect benefits of each strategy, which might also differ and vary in relation to the evolution of the other strategy. Dynamic models are consequently required to understand and predict evolutionary outcomes.

We address three key questions for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given strong inbreeding depression. These questions include: (1) Does evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance facilitate evolution of polyandry that is otherwise severely constrained from increasing relatedness of additional mates? (2) How do costs associated with polyandry and pre-copulatory and post-copulatory inbreeding avoidance affect evolutionary outcomes and, in particular, the long-term persistence of reproductive strategies given cost asymmetry? Finally, (3) how is selection for initially rare pre-copulatory or post-copulatory inbreeding avoidance affected if the other strategy of inbreeding avoidance is already fixed? By addressing these questions, we provide a framework for exploring the simultaneous evolution of multiple reproductive strategies through the use of individidual-based modelling, and we derive hypotheses amenable to empirical tests within and across taxa.

>>>>

Developing theory on evolution of polyandry given inbreeding depression, @Duthie used a genetically-explicit individual-based model to examine the conditions under which polyandry is predicted to evolve solely due to selection stemming from pre-copulatory inbreeding avoidance. A recent genetically-explicit individual-based model examined the conditions under which polyandry is predicted to evolve solely due to selection stemming from pre-copulatory inbreeding avoidance [@Duthie]. Evolution of polyandry occurred only under highly restricted conditions, requiring that direct negative selection (i.e., 'costs') on polyandry were small, that very few males were available for a female's initial mate choice but many were available for additional mate choice(s), or that polyandry was conditionally expressed when a focal female was related to her initial mate [@Duthie]. Without these conditions, increasing polyandry tended to increase rather than decrease the overall degree of inbreeding that females experienced, and hence reduce offspring fitness. This occurred because, once pre-copulatory inbreeding avoidance evolved, polyandrous females had already chosen available unrelated males as their initial mates. Their additional mates, chosen from the remaining male population, were therefore increasingly likely to include relatives. Evolution of polyandry purely to facilitate pre-copulatory inbreeding avoidance, and hence the resulting degree of inbreeding avoidance, was consequently restricted [@Duthie].

<!--- I don't think 'provide initial theory' is particularly clear (i.e., 'initial' in what respect?), so I've made an edit to the first sentence, though it does seem slightly redundant in both of our cases. --->

However, if post-copulatory inbreeding avoidance could also evolve, polyandrous females would also have an opportunity to bias paternity towards unrelated males within their set of chosen mates. Evolution of post-copulatory inbreeding avoidance might consequently reduce the cost of polyandry stemming from the accumulation of related mates across multiple matings, potentially facilitating evolution of polyandry to avoid inbreeding under broader conditions, and driving further evolution of pre-copulatory or post-copulatory mate choice strategies. However, if polyandry and pre-copulatory and post-copulatory inbreeding avoidance can all evolve, the long-term dynamics of these three reproductive strategies become difficult to predict. Strong inbreeding depression might drive initial evolution of both pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. However, the co-occurence of pre-copulatory and post-copulatory inbreeding avoidance might engender some degree of `degeneracy', defined as a phenomenon by which different elements of a system result in identical outputs [i.e., @Edelman2001]. Consequently, if evolution of polyandry and post-copulatory inbreeding avoidance renders pre-copulatory inbreeding avoidance functionally redundant, or vice versa, then only one inbreeding avoidance strategy might be maintained in the long-term.

<!--- the 'polyandry and pre-copulatory and post-copulatory inbreeding avoidance' seems technically off. It relies on readers to recognise that 'inbreeding avoidance' only applies to the first two items in the list. I suspect this shouldn't be a problem in practice, but I wonder if there's a better way to word this  --->

<!--- Regarding the removal of 'conceptualise', are we suggesting then that prediction is only or most important thing to predict in the long term, and that we can do it with our model? I'm not sure if I agree with either. --->

@@

The few previous models that considered the evolution of inbreeding avoidance through mate choice (as opposed to dispersal) have implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models are useful for isolating the invasion fitness of this single strategy. However, when both pre-copulatory and post-copulatory strategies can affect the realised degree of inbreeding, it cannot be assumed that both strategies will simultaneously invade a randomly mating population, nor that the invasion fitness of one strategy will be independent of the pre-existence or invasion fitness of the other strategy. For example, if pre-adaptation or a selective sweep results in fixation of alleles underlying pre-copulatory inbreeding avoidance, then new alleles underlying polyandry and post-copulatory inbreeding avoidance might be unlikely to invade a population because the phenotypic effect of such alleles on the overall degree of inbreeding, and resulting indirect selection, could be negligible. Conversely, fixation of alleles underlying polyandry and post-copulatory inbreeding avoidance might reduce or eliminate positive selection on alleles underlying pre-copulatory inbreeding avoidance and hence impede adaptive evolution of mate choice. New theory, guided by modelling that evaluates invasion dynamics of alleles underlying multiple interacting and potentially functionally redundant (i.e., `degenerate') traits, is therefore needed.

<!--- I'm not entirely sure the 'resulting indirect selection' helps --->

<!--- I've made the suggested hedging changes. I originally introduced things boldly with 'preclude' or 'eliminate' by themselves as a way to introduce readers to the idea of the logic in the strongest terms possible, without thinking that readers would try to draw definite conclusions in the introduction or dwell on exceptions. Of course, as with anything in biology, the empirical (or even IBM) reality is going to be messier.  --->

<!--- Along the lines of the above then, I don't think it's always helpful to use 'evolution' as a short-hand for when we mean 'adaptive evolution'. For example, elimation of positive selection on alleles might not 'impede evolution' at all (which will still happen as long as there is mutation, drift, etc.), but rather change the direction of evolution. --->

<!--- I don't understand why 'the simultaneous' invasion dynamics was being emphasised. Isn't the point of invasion analysis to look at one thing at a time and see whether or not the trait can invade (this is what we talk about in this paragraph, at least, and what we do later). By constraints, I mean something to the effect of *not having everything start out at zero and evolve simultaneously, which is biologically unrealistic*.--->

<!--- I still don't like this last sentence very much. It doesn't really explain *why* new theory is needed, nor why such a specific set of modelling assumptions are required to address the theoretical question. I'm not even sure that I see a clear theoretical question in there. --->

Ultimately, the absolute and relative frequencies of alleles underlying pre-copulatory and post-copulatory inbreeding avoidance and polyandry will depend not only on the magnitudes of positive indirect selection stemming from reduced inbreeding depression in females' offspring, but also on the magnitudes of direct negative selection on resulting phenotypes (i.e., the direct fitness costs of expressing each reproductive strategy). Polyandrous females can pay costs of mate searching or mating, for example, because these activities increase predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice and consequent reluctance to mate with relatives might pay costs of increased risk of complete mating failure or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay energetic costs associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs of pre-copulatory and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly strategy might become fixed over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy. However, such trajectories will also depend on the indirect benefits of each strategy, which might also differ and vary in relation to the evolution of the other strategy. Dynamic models that track the frequencies of alleles underlying multiple reproductive strategies that might influence each other's evolution across generations, and that are enacted among relatives that themselves result from reproductive strategies and inbreeding depression expressed in previous generations, are consequently useful to understand and predict evolutionary outcomes.

<!--- Do we really need ''absolute and relative'' frequencies? This just gets me thinking about the distinction and why it might be important, but we don't actually talk about that here. --->

<!--- I think that the addition to the last sentence here is probably an improvement, but it still doesn't quite address my question -- there's not much of an argument for concluding that dynamic models are *required*. The implicit argument seems to be something like, ''look, there's a lot of complex stuff that's happening here; we can't figure out how to model how the evolutionary trajectories will play out without a dynamic model, so we're just going to assume that it's not possible''. My tendency is to be careful with these kinds of things, and maybe instead emphasise that dynamic models *can be useful* for understanding these kinds of dynamics without suggesting that they are necessary (time will tell if this is true). --->

We use individual-based modelling to address three key questions regarding evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given strong inbreeding depression. First, does evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance facilitate evolution of polyandry that is otherwise severely constrained from increasing relatedness of additional mates? Second, how do costs associated with polyandry and pre-copulatory and post-copulatory inbreeding avoidance affect evolutionary outcomes and, in particular, the long-term persistence of reproductive strategies given cost asymmetry? Third, how is selection for initially rare pre-copulatory or post-copulatory inbreeding avoidance affected if the other strategy of inbreeding avoidance is already fixed? By addressing these questions, we illustrate how the simultaneous evolution of multiple interacting degenerate reproductive strategies can generate diverse evolutionary outcomes; we thereby provide a framework for deriving hypotheses amenable to empirical tests within and across taxa.

<!--- the 'outcomes; we' replaces the 'outcomes and' just to avoid making it sound like the second part of the sentence follows just from 'diverse evolutionary outcomes' instead of the whole first part. --->
